---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = "center", 
  out.width = "100%",
  fig.width = 9, fig.height = 7
)
```

<!-- badges: start -->
<!-- badges: end -->


We introduce the `NAIR` package, a powerful tool for analyzing the adaptive immune repertoire using network analysis based on similarities among receptor sequences, which implements methods from the following paper:

[Hai Yang, Jason Cham, Brian Neal, Zenghua Fan, Tao He and Li Zhang. (2023). NAIR: Network Analysis of Immune Repertoire. *Frontiers in Immunology*, vol. 14. https://doi.org/10.3389/fimmu.2023.1181825](https://www.frontiersin.org/articles/10.3389/fimmu.2023.1181825)

Our immune repertoire analysis tool allows users to perform various network analysis tasks on AIRR-Seq data. This includes computing local and global network properties of nodes and clusters, which can provide insights into the structural organization of the immune repertoire network.

In addition, our tool enables users to search across multiple AIRR-Seq samples for clones/clusters associated with subject characteristics, disease conditions or clinical outcomes, as well as identify public clones/clusters. This can help researchers identify potentially important TCR/BCR clones.

To aid in visualization and interpretation of the immune repertoire network, our tool also allows for customized visualizations of the network. Furthermore, our tool enables users to perform further downstream analysis on the immune repertoire data, such as clustering analysis and differential abundance testing. These downstream analyses can provide additional insights into the immune response.


### What data does `NAIR` support?

`NAIR` supports bulk and single-cell immune repertoire sequence data for T-cell or B-cell receptors (TCR or BCR).

* **Single-cell data:** Each row is a single cell
* **Bulk data:** Each row is a distinct TCR/BCR clone (unique combination of V-D-J genes and nucleotide sequence) and typically includes a corresponding measurement of clonal abundance (e.g., clone count and clone frequency/fraction)


### How does `NAIR` model the immune repertoire as a network?

* Each cell (single-cell data) or clone (bulk data) is modeled as a node (vertex) in the network
* For each node, we consider the corresponding receptor sequence (nucleotide or amino acid)
* For each pair of nodes, we measure the similarity in their receptor sequences (using the Hamming or Levenshtein distance)
* An edge is drawn between two nodes if the distance is below a specified threshold 
    * For single-cell data, sequences from two chains (e.g., alpha chain and beta chain) can be jointly used to determine similarity between cells, considering cells as similar when the sequences for both chains are similar (i.e., when the distance for each chain is below the threshold)
* Clustering analysis is used to partition the network graph into clusters (densely-connected subgraphs)
    * Many clustering algorithms are available, with each seeking to identify the "best" configuration of clusters according to different graph criteria
* Network statistics characterize the repertoire in terms of the local and global structural properties of its graph
* Customized visual plots of the network graph are generated, with nodes colored according to desired meta-data (e.g., disease status, sample ID, cluster ID, clone count, etc.)



# Installation

The current development version of `NAIR` can be installed from github using the following commands:

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github(
  "mlizhangx/Network-Analysis-for-Repertoire-Sequencing-",
  build_vignettes = TRUE)
```

Installing the development version requires a toolchain compiler. On Windows, this means downloading and installing Rtools. On MacOS, this entails installing XCode Command Line Tools ("XCode CLI") and the correct version of gfortran for your macOS version (instructions [here](https://thecoatlessprofessor.com/programming/cpp/r-compiler-tools-for-rcpp-on-macos/)).




# Documentation

Once the package is installed, type the following commands to access the package vignettes and documentation:

```{r, eval = FALSE}
# Package introduction vignette
vignette("NAIR", package = "NAIR")

# Display vignettes in html browser
browseVignettes("NAIR")

# Browse documentation
help(package = "NAIR")
```





# The `buildRepSeqNetwork` function

General network analysis on AIRR-Seq data is performed using the `buildRepSeqNetwork` function. This function does the following:

* Filters the input AIRR-Seq data according to user specifications
* Builds the network graph for the immune repertoire
* Performs desired network analysis
    * Cluster analysis
    * Computation of network properties
* Generates customized visual plot of the network graph using `ggraph`
* Saves and returns the following output: 
    * Meta-data for the nodes in the network, including network properties
    * Meta-data for the clusters in the network
    * Network graph plot (returned as `ggraph` object and saved as pdf)
    * Network adjacency matrix in sparse matrix format
    * `igraph` object containing the list of edges in the network graph 

See `vignette("buildRepSeqNetwork")` for a tutorial on the function's usage and output.



# Visualization

The `buildRepSeqNetwork` function includes various arguments that facilitate customization of the network visualization. See `vignette("network_visualization")` for an overview.




# Searching for Associated Clusters

Given multiple samples of AIRR-Seq data, the `NAIR` package can be used to search for TCR/BCR clusters associated with a binary variable of interest, such as a disease condition, treatment or clinical outcome. 

See `vignette("associated_clusters")` for a detailed tutorial.



# Searching for Public Clusters

The `NAIR` package includes a set of functions that facilitate searching for public TCR/BCR clusters across multiple samples of AIRR-seq data. 

In this context, a public cluster consists of similar TCR/BCR clones (e.g., those whose CDR3 amino acid sequences differ by at most one amino acid) that are shared across samples (e.g., across individuals or across time points for a single individual). 

See `vignette("public_clusters")` for a detailed tutorial.
