---
title: "Searching for Associated TCR/BCR Clusters"
output: 
  rmarkdown::html_vignette:
    number_sections: false
    tabset: true
vignette: >
  %\VignetteIndexEntry{Searching for Associated TCR/BCR Clusters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center", 
  out.width = "100%",
  fig.width = 9, fig.height = 7
)
```

# Introduction

Given multiple samples of Adaptive Immune Receptor Repertoire Sequencing (AIRR-Seq) data, the `NAIR` package can be used to search for TCR/BCR clusters associated with a binary variable of interest, such as a disease condition, treatment or clinical outcome. 

We first provide a brief conceptual overview, followed by a demonstration in which we explain the process in greater detail.


## Overview of Process

1. **Identify associated sequences.** Divide the subjects into two groups based on the two levels of the binary variable. Identify TCR/BCR sequences that exhibit a statistically significant difference in frequency between the two groups using Fisher's exact test.
2. **Identify clones with sequences similar to the associated sequences.** For each associated sequence, all sequences that fall within a certain distance (e.g., those that differ by at most one amino acid) comprise its neighborhood. From all samples, identify all clones whose sequences belong to this neighborhood.
3. **Construct global network using identified clones and perform clustering.** Combine the clones from all neighborhoods into a single global network. Perform cluster analysis and assign membership to the global clusters. These clusters are considered as the associated clusters.


## Simulate Data for Demonstration

We simulate some toy data for demonstration.

As our binary variable of interest, we consider a single treatment factor with two levels, labeled treatment and control. 

In each of the two groups, we simulate 15 samples, each containing 30 observations. The generation probabilities of the possible sequences are fixed within each group. In order to simulate the treatment effect, the generation probabilities of certain sequences differ substantially between the two groups.

Each sample is saved in a separate file using the .rds file format. The files are named "`Sample1.rds`", "`Sample2.rds`", etc. The file path of their directory is saved to the R environment variable `dir_input_samples` for later reference.

```{r }
# Use temp dir
data_dir <- tempdir()

# Directory to store input files
dir_input_samples <- file.path(data_dir, "input_samples")
dir.create(dir_input_samples, showWarnings = FALSE)

# Number of samples by control/treatment group
samples_c <- samples_t <- 15 
samples <- samples_c + samples_t
sample_size <- 30 # (seqs per sample)          

# sequences (first five are chosen to be associated with treatment)
base_seqs <- c("CASSGAYEQYF", "CSVDLGKGNNEQFF", 
               "CASSIEGQLSTDTQYF", 
               "CASSEEGQLSTDTQYF",
               "CASSPEGQLSTDTQYF",
               "RASSLAGNTEAFF", "CASSHRGTDTQYF", "CASDAGVFQPQHF") 

# relative generation probabilities by control/treatment group
pgen_c <- matrix(rep(c(rep(1, 5), rep(30, 3)), times = samples_c), 
                 nrow = samples_c, byrow = TRUE)
pgen_t <- matrix(rep(c(1, 1, rep(1/3, 3), rep(2, 3)), times = samples_t), 
                 nrow = samples_t, byrow = TRUE)
pgen <- rbind(pgen_c, pgen_t)

# Simulate the data
library(NAIR)
simulateToyData(    
  samples = samples, sample_size = sample_size,
  prefix_length = 1, prefix_chars = c("", ""),
  prefix_probs = cbind(rep(1, samples), rep(0, samples)),
  affixes = base_seqs, affix_probs = pgen, num_edits = 0,
  output_dir = dir_input_samples, no_return = TRUE)
```

The first few rows of the data for the first sample appear as follows:

```{r}
# View first few rows of data for sample 1
head(readRDS(file.path(dir_input_samples, "Sample1.rds")))
```


# <a name="step1"></a> Step 1: Find Associated Sequences {.tabset}

The first step is to conduct a systematic search for associated sequences within the provided samples using the `findAssociatedSeqs` function. 

This search is a two-stage procedure. 

The unique receptor sequences are first filtered according to basic criteria in order to narrow the list of candidates. 

For each candidate sequence, we then compute the P-value for Fisher's exact test of independence between the binary variable of interest and the observed presence of the sequence within a sample/subject. The user specifies the P-value cutoff below which an association is detected.

Below, we explain the usage and behavior of the `findAssociatedSeqs` function.


## Search Criteria {.tabset}

The `findAssociatedSeqs` function has several parameters that control the criteria used when searching for associated sequences. These include several filter criteria used to determine which sequences are considered for testing, as well as an argument to specify the P-value cutoff below which an association is detected. These arguments are presented below.

### Sample Membership

By default, only sequences that appear in at least five samples will be considered. This can be changed by setting the `min_sample_membership` argument to a different value. Setting the value to `NULL` bypasses this check.

### Sequence Length

By default, only sequences that contain at least seven characters will be considered. This can be changed by setting the `min_seq_length` argument to a different value. Setting the value to `NULL` bypasses this check.

### Sequence Content

Sequences containing characters `*` or `_` will be excluded from consideration. This can be changed using the `drop_matches` argument, which accommodates a character string or regular expression specifying the pattern of content to search for. The content of each sequence is checked for a match to this pattern using the `grep` function from base R. If a match is found, the sequence is excluded from consideration. Setting the value to `NULL` bypasses this check. 

For details on how the pattern matching is performed, please refer to the base R documentation files for `regex` and `grep`.

### P-value Cutoff

By default, sequences with a P-value below 0.05 on Fisher's exact test are included in the output of `findAssociatedSeqs`. However, users may wish to impose a stronger burden of evidence to control the false discovery rate under multiple testing. The cutoff can be set to a different value using the `pval_cutoff` argument. The lower the cutoff value, the stronger the evidence of an association is required for a sequence to be included in the output. 

It should be noted, however, that the sequences returned by `findAssociatedSeqs` are ordered by P-value, and any subset of them can be used in [step2](#step2). Thus, imposing a stricter P-value cutoff can also be done indirectly after calling `findAssociatedSeqs`, by subsetting the results according to P-value.


## <a name="input_files"></a> Input File Paths

The main argument of the `findAssociatedSeqs` function is the `file_list` argument, which accepts a vector containing file paths. Each path corresponds to a distinct AIRR-Seq data file representing an individual sample.

Below, we prepare the vector `input_files` to be provided to the `file_list` argument of `findAssociatedSeqs`:

```{r}
# input files for step 1 (one per sample)
input_files <- file.path(dir_input_samples, paste0("Sample", 1:samples, ".rds"))
head(input_files)
```

## <a name="input_type"></a> Input File Format

The file format of the input files for `findAssociatedSeqs` is specified using the `input_type` parameter. The supported formats include `"rds"`, `"rda"`, `"csv"`, as well as files that can be read using the `read.table` function, such as `"tsv"` and `"txt"`. 

For text formats such as `"csv"`, `"tsv"` and `"txt"`, users can specify the separation option by utilizing the `sep` argument. The default setting `sep = ""` accommodates all forms of white space, i.e., one or more spaces, tabs, newlines or carriage returns. In addition, it is important to note that the first line of the data is assumed to be the header by default. To disable this behavior and treat the first line as data, users must set the `header` parameter to `FALSE`.

For the `"rda"` (Rdata) format, the `data_symbols` argument is used to specify the name of each sample's AIRR-Seq data frame within the Rdata file (i.e., the name of the data frame as it will appear in the R environment when the Rdata file is loaded). The argument accepts a character string. This requires that each sample's data frame has the same name within its respective Rdata file. For example, if each sample's data frame is named `mydataframe` and saved using the command `save(mydataframe, file = myfilepath)`, where the file path `myfilepath` is different for each sample, then we would set `data_symbols = "mydataframe"`.

Our samples are stored in .rds files, so we use `input_type = "rds"`. 


## <a name="seq_col"></a> Specifying the Sequence Column

The `seq_col` argument is used to specify the column containing the clone sequences in the input data for each sample. The argument accepts either the column name or column index.

In our simulated data, the column containing the clone sequences is named `CloneSeq`.


## <a name="group_ids"></a> Assigning Samples to Groups

The `group_ids` argument is used to assign each sample to one of the two groups, representing the two levels of the binary variable of interest. The argument accepts a vector of the same length as `file_list`. Each entry of `group_ids` is assigned as a group label to the sample in the corresponding entry of `file_list`. Any values may be used for the group labels, but the vector must contain exactly two unique values.

For instance, in our simulated data, the first half of the samples belong to the control group, while the second half belong to the treatment group. Thus, we should assign one group label to the first 15 samples and a different group label to the last 15 samples. Here we choose to label the first 15 samples as `"reference"` and the last 15 samples as `"comparison"`. However, it should be noted that the results will be unchanged if the labels are reversed, or if a different pair of labels is used, as long as the first 15 samples are assigned to one group and the last 15 samples to the other.

The vector we will provide to the `group_ids` argument of `findAssociatedSeqs` is created below:

```{r}
# group label assignments for the samples
group_labels <- c(rep("reference", samples_c), rep("comparison", samples_t))
group_labels
```


## <a name="subject_ids"></a> Assigning Samples to Subjects (If Applicable)

The `subject_ids` argument can be used to assign each sample to a particular subject. The argument accepts a vector of the same length as `file_list`. Each entry of `subject_ids` is assigned as a subject ID to the sample in the corresponding entry of `file_list`.

If the `subject_ids` argument is omitted, Fisher's exact test treats each sample as an independent observational unit. In this case, the relevant contingency table involves counts of **samples** possessing a given sequence.

If subject IDs are provided, each subject's collection of samples is treated as a single observational unit. The relevant contingency table then involves counts of **subjects** possessing a given sequence. This allows a sequence to be counted at most once per subject, and results in each subject being counted exactly once in each margin.

Subject IDs should be provided when the binary variable of interest is subject-specific and the data contains multiple samples from a single subject. Subject IDs should be omitted when the binary variable of interest is sample-specific or each sample comes from a different subject.



## <a name="freq_col"></a> Specifying the Clone Frequency Column (Optional)

The `freq_col` argument can be used to specify a column containing clone frequencies in the input data for each sample. The argument accepts either the column name or column index.

If clone frequencies are provided, the maximum clone frequency (across all samples) for each associated sequence will be included in the content of the data frame returned by `findAssociatedSeqs`.

## Output File Path

The `findAssociatedSeqs` function returns a data frame containing the associated sequences along with some additional information. The format and contents of this data frame will be explained below after executing the function.

By default, the data frame returned by `findAssociatedSeqs` is also saved to the current working directory as a csv file named `associated_seqs.csv`.

A different file name and/or directory can be specified by providing a file path to the `outfile` argument. For example, setting `outfile = "myfile.csv"` will save the file to the current working directory as `myfile.csv`, while setting `outfile = "~/myfolder/myfile.csv"` will save the file within the subdirectory `myfolder` located within the current working directory.

The user can also specify `outfile = NULL` in order to prevent the output from being saved.


## Demonstration of Execution and Output

We execute the `findAssociatedSeqs` function using the inputs we prepared earlier for the `file_list` and `group_ids` arguments:

```{r}
# search across samples for associated sequences using Fisher's exact test
associated_seqs <- findAssociatedSeqs(
  file_list = input_files, input_type = "rds", 
  group_ids = group_labels, 
  seq_col = "CloneSeq", 
  min_seq_length = NULL, drop_matches = NULL, 
  min_sample_membership = NULL, 
  pval_cutoff = 0.1,
  outfile = NULL)
```

`findAssociatedSeqs` returns a data frame containing the receptor sequences found to be associated with the binary variable based on Fisher's exact test using the specified P-value cutoff. Each row corresponds to a unique sequence and includes the following variables: 

* `ReceptorSeq`: The unique receptor sequence
* `fisher_pvalue`: The P-value on Fisher's exact test for independence between the receptor sequence and the binary variable of interest
* `shared_by_n_samples`: The number of samples in which the sequence was observed
* `samples_g0`: Of the samples in which the sequence was observed, the number of samples belonging to the first group (first unique value of `group_ids`) 
* `samples_g1`: Of the samples in which the sequence was observed, the number of samples belonging to the second group (second unique value of `group_ids`) 
* `shared_by_n_subjects`: The number of subjects in which the sequence was observed (only present if subject IDs are specified through `subject_ids`)
* `subjects_g0`: Of the subjects in which the sequence was observed, the number of subjects belonging to the first group (only present if subject IDs are specified through `subject_ids`)
* `subjects_g1`: Of the subjects in which the sequence was observed, the number of subjects belonging to the second group (only present if subject IDs are specified through `subject_ids`)
* `label`: A character string summarizing the above information. Also includes the maximum in-sample clone frequency across all samples, if available.

```{r}
# view first few rows of output
head(associated_seqs)
```

The rows of the data frame are ordered by Fisher's exact test $P$-value.


# <a name="step2"></a> Step 2: Find Associated Clones {.tabset}

In the [previous step](#step1), we used `findAssociatedSeqs` to identify receptor sequences associated with the binary variable of interest.

The next step is to use the `findAssociatedClones` function to search across samples and identify all clones with sequences similar to the associated sequences identified in step 1.

For each associated sequence, we define its neighborhood as the set of all sequences that fall within a specified distance (e.g., a maximum Hamming distance of 1). We then identify all clones (from all samples) with sequences belong to this neighborhood.

The data for each associated sequence's neighborhood is then saved to a separate file to be used as an input in [step 3](#step3).


## Specifying the Sample Data

In order for `findAssociatedClones` to conduct its search, we must specify our sample data, just as we did when calling `findAssociatedSeqs` in step 1. 

This is done using the arguments  `file_list`, `input_type`, `group_ids` and `seq_col`, which behave in the same manner as seen earlier for the `findAssociatedSeqs` function in [step 1](#step1):

* [`file_list`](#file_list) accepts a vector of input file paths
* [`input_type`](#input_type) accepts a string specifying the input file format
    * `sep`, `header` and `data_symbols` are situational arguments regarding data format
* [`group_ids`](#group_ids) accepts a vector of group assignments
* [`seq_col`](#seq_col) accepts the name or index of the clone sequence data column

Refer back to [step 1](#step1) for a detailed explanation of these arguments.

The user should use the same values for these arguments as they did when calling `findAssociatedSeqs` in [step 1](#step1).


## Assigning Custom Sample IDs (Optional)

Each clone's sample ID is automatically included as a variable in the data for each associated sequence's neighborhood. 

By default, the samples are labeled numerically according to the order they appear in `file_list`. 

The `sample_ids` argument allows for custom sample IDs to be assigned. It accepts a vector of the same length as `file_list`. Each entry of `sample_ids` is assigned as a sample ID to the sample in the corresponding entry of `file_list`. 



## Assigning Subject IDs (Optional)

The `subject_ids` argument allows for a subject ID to be assigned to each sample as a separate variable from sample ID. 

If subject IDs are provided, each clone's subject ID will be included as a variable in the data for each associated sequence's neighborhood (in addition to the sample ID variable). 

The `subject_ids` argument accepts a vector of the same length as `file_list`. Each entry of `subject_ids`is assigned as a subject ID to the sample in the corresponding entry of `file_list`. 


## Specifying the Associated Sequences

The associated sequences are specified via the `assoc_seqs` argument, which accepts a character vector containing the associated sequences.

Typically, the vector provided to `assoc_seqs` will be the `ReceptorSeq` column of the data frame returned by `findAssociatedSeqs`. This considers all of the associated sequences found in step 1. 

However, it may be desirable to consider only a subset of the sequences returned by `findAssociatedSeqs`. The sequences are ordered by Fisher's exact test P-value to facilitate reference. For example, if we had many associated sequences, we could choose to consider only the 10 with the lowest P-values by specifying `assoc_seqs = associated_seqs$ReceptorSeq[1:10]`.


## Neighborhood Distance Settings

By default, each associated sequence's neighborhood includes all sequences with a Hamming distance of at most 1 from the associated sequence.

The type of distance metric and the distance threshold used to determine the neighborhoods can be adjusted using the `dist_type` and `nbd_radius` arguments. 

For example, setting `dist_type = "lev"` and `nbd_radius = 2` results in each neighborhood containing all sequences with a Levenshtein distance of at most 2 from the associated sequence.


## Output Settings {.tabset}

The `findAssociatedClones` function does not return any direct output. Instead, it saves the data for the associated sequence's neighborhoods to files that will be used as inputs in [step 3](#step3). 

For each neighborhood, a data frame is saved to an individual file. The data frame contains one row for each clone in the neighborhood. The following arguments can be used to customize the output. 

### Subsetting Columns From the Input Data

The data frame for each neighborhood includes all of the data columns that were present in each sample's original input data frame. 

In some cases, the user may not wish for all columns to be included. In this case, the user can specify the columns to be included using the `subset_cols` argument, which accepts either a vector of column names or a vector of column indices. 

The column specified by the `seq_col` argument will automatically be included regardless of the value of `subset_cols`.

### Output Directory

The file path for the output directory is specified using the `output_dir` argument. The output directory specified by the `output_dir` argument will be created if it does not already exist. 

```{r}
# output directory for current step
dir_nbds <- file.path(data_dir, "assoc_seq_nbds")
```

### Output File Format

By default, each neighborhood's data frame is saved as a csv file. This can be changed using the `output_type` argument. Other valid options include `"tsv"`, `"rds"` and `"rda"`. 

If files are saved using the `"rda"` (Rdata) format, each data frame will be named `data` in the R environment when its Rdata file is loaded. 


## Example Function Call

We execute the `findAssociatedClones` function using the inputs we prepared earlier for the `file_list` and `group_ids` arguments:

```{r}
# Identify clones in a neighborhood around each associated sequence
findAssociatedClones(
  file_list = input_files, input_type = "rds", 
  group_ids = group_labels, 
  seq_col = "CloneSeq", 
  assoc_seqs = associated_seqs$ReceptorSeq,
  min_seq_length = NULL, drop_matches = NULL,
  output_dir = dir_nbds)
```

The following files are created in the directory specified by the `output_dir` argument:

```{r}
# Files created by findAssociatedClones
list.files(dir_nbds)
```

Each file contains the neighborhood data for a single associated sequence.


# <a name="step3"></a> Step 3: Global Network of Associated Clusters {.tabset}

Now that we have identified the clones in each associated sequence's neighborhood, the final step is to use the `buildAssociatedClusterNetwork` function to combine the clones from all neighborhoods into a single global network. We then use clustering analysis to partition the global network into clusters, which are considered as the associated clusters.


## Input File Paths

The files created by `findAssociatedClones` in the [previous step](#step2) contain the data for each neighborhood. 

These files are provided to `buildAssociatedClusterNetwork` by supplying a character vector of file paths to the `file_list` argument. We create this vector below.

```{r}
# Files created by findAssociatedClones
nbd_files <- list.files(dir_nbds, full.names = TRUE)
```


## Input File Format 


By default, `findPublicClusters` saves its output files using the RDS file format. By default,  `buildPublicClusterNetwork` assumes that the input files are in the RDS file format. 

If the files created by `findPublicClusters` were saved using a different file format, the `input_type` argument can be used to specify the format used. This argument has the same behavior as seen in `findPublicClusters`.

For the `"rda"` (Rdata) file format, the `data_symbols` argument is used to specify the name of each sample's node-level data frame within its respective Rdata file. If `findPublicClusters` was called with `output_type = "rda"`, then the node-level data frame for each sample will be named `ndat`. Accordingly, the default value of `data_symbols` is `ndat`, so the user should not need to manually specify the value of `data_symbols` unless the names within the files are changed.



## Customization of Network Analysis {.tabset}

`buildAssociatedClusterNetwork` uses the same arguments as [`buildRepSeqNetwork`](buildRepSeqNetwork.html) for customizing the global network analysis. [This vignette](buildRepSeqNetwork.html) provides an overview of these arguments. We mention some of the important ones below.


### Distance Function and Distance Cutoff

By default, two nodes within the global network are joined by an edge if the Hamming distance between their sequences is at most 1. 

The Levenshtein distance can be used instead of the Hamming distance by setting `dist_type = "lev"`. 

The maximum distance for two nodes to be joined by an edge can be changed using the `dist_cutoff` argument.


### Clustering Algorithm

After constructing the global network, `buildAssociatedClusterNetwork` performs cluster analysis on the network nodes, partitioning the global network graph into densely-connected subgraphs. 

If the neighborhoods for two different associated sequences are very similar, i.e., the network contains many edge connections joining the nodes from one neighborhood to the nodes from the other neighborhood, then the two neighborhoods will have a high chance of belonging to the same cluster. Thus, cluster analysis assists in identifying distinct groups of clones/sequences that are associated with the binary variable of interest.

By default, clustering is performed using the `cluster_fast_greedy` algorithm from the `igraph` package. Other clustering algorithms from the `igraph` package can be used instead of the default algorithm. The algorithm is specified using the `cluster_fun` argument, which accepts one of the following functions:

* `cluster_fast_greedy` 
* `cluster_edge_betweenness` 
* `cluster_fluid_communities` 
* `cluster_infomap` 
* `cluster_label_prop` 
* `cluster_leading_eigen` 
* `cluster_leiden` 
* `cluster_louvain` 
* `cluster_optimal` 
* `cluster_spinglass` 
* `cluster_walktrap`

For example, setting `cluster_fun = cluster_leiden` performs clustering using the `cluster_leiden` algorithm.

For more information about a particular algorithm, users can refer to its help documentation file. For example, the command `?cluster_fast_greedy` loads the documentation file for the `cluster_fast_greedy` (the default) algorithm, assuming the `NAIR` package has been loaded (e.g., using `library(NAIR)`).


## Customization of Visual Plot

By default, the network graph plot produced by `buildAssociatedClusterNetwork` colors the nodes according to the binary variable of interest. This assists the user in validating each cluster's association to the binary variable. It also allows one to visually distinguish clusters in which only a single level of the binary variable is present (e.g., disease-only clusters). 

If desired, a different variable can be used to color the nodes. This is done by specifying the column for the variable to the `color_nodes_by` argument, which accepts a column name or column index. 

The `color_nodes_by` argument also accommodates a vector of column names or a vector of column indices, in which case one plot will be created for each column specified, with each plot coloring the nodes according to the variable in its respective column. For example, setting `color_nodes_by = c(3, 5)`  will produce two plots per sample, where one plot has the nodes colored according to the value in the third column of the input data

[This vignette](network_visualization.html) details additional arguments that can be used to further customize the visual plot. 


## Output Settings

The output returned by `buildAssociatedClusterNetwork` follows the same format as the output of [`buildRepSeqNetwork`](buildRepSeqNetwork.html). The function returns a list containing the node-level and cluster-level meta data for the global network, as well as any plots generated, in addition to the network adjacency matrix and the `igraph` network edge list.

By default, the contents of the list returned by `buildAssociatedClusterNetwork` are saved to the current working directory. Each list element is saved as an individual file. The file formats are the same default file formats used by [`buildRepSeqNetwork`](buildRepSeqNetwork.html). In particular, the node-level and cluster-level meta data are saved as csv files. 

Alternatively, the user can save the entire output list to a single compressed rds or rda file by setting `output_type = "rds"` or `output_type = "rda"`, respectively.

By default, all files saved share the common file name prefix `"AssociatedClusterNetwork"`. This common file name prefix can be set to a different value by supplying a character string to the `output_name` argument.

The output can be saved to a different directory by providing a file path to the `output_dir` argument. 

The user can also specify `output_dir = NULL` in order to prevent the output from being saved.



## Execution and Output

We execute the `buildAssociatedClusterNetwork` function using the input we prepared earlier for the `file_list` argument:

```{r}
# Combine neighborhoods and perform network analysis
all_clusters <- buildAssociatedClusterNetwork(
  file_list = nbd_files, 
  seq_col = "CloneSeq", 
  size_nodes_by = 1.5,
  output_dir = file.path(data_dir, "assoc_clusters"))
```

In the visual plot of the network, we see that the treatment group shows high representation in the clusters, affirming the association between these sequences and the treatment variable.

The elements of the output list returned by `buildAssociatedClusterNetwork` have the following names:

```{r}
# output returned by buildAssociatedClusterNetwork
names(all_clusters)
```

The default files saved by `buildAssociatedClusterNetwork` have the following names:

```{r}
# Files saved by buildAssociatedClusterNetwork
list.files(file.path(data_dir, "assoc_clusters"))
```

### Node-Level Meta Data

The `node_data` data frame contained in the output list contains the following variables:

```{r}
# variables in the node-level meta data
names(all_clusters$node_data)
```

Notice that by default, all variables that were present in each sample's original data, such as `"CloneFrequency"` and `"CloneCount"`, are automatically carried over into this data.

In addition, variables containing various node-level network properties are also present.


### Cluster-Level Meta Data

The `cluster_data` data frame contained in the output list contains the following variables:

```{r}
# variables in the node-level meta data
names(all_clusters$cluster_data)
```

Each row corresponds to a cluster in the global network, and each variable corresponds to a cluster-level property.


## Labeling the Global Clusters

In order to more easily cross-reference the clusters in the visual plot with the clusters in the data, we can [label the clusters with their ID numbers](network_visualization.html#labeling-clusters) as follows:

```{r}
# Modify plot to add labels to the clusters
all_clusters$plots[[1]] <- 
  addClusterLabels(
    plot = all_clusters$plots[[1]],
    net = all_clusters,
    top_n_clusters = 3,
    criterion = "node_count", # (the default)
    size = 10
  )

# View modified plot
all_clusters$plots[[1]]
```


## Focusing on Individual Clusters of Interest

If we wish to focus on a particular cluster of interest within the global network, we can build a network exclusively using the clones from that cluster. 

Below, we focus on the first cluster, which is also the largest cluster by node count.

We generate two plots. In the first plot, we color each node according to the receptor sequence of its corresponding clone. This provides a more detailed account of the sequences that appear in the cluster and their relative representation.

In the second plot, we color each node according to the sample in which the corresponding clone originally appeared. This allows one to distinguish whether the clones in the cluster come from many samples as opposed to relatively few samples.

```{r}
# focus on the first cluster
buildRepSeqNetwork(
  data = all_clusters$node_data[all_clusters$node_data$cluster_id == 1, ],
  seq_col = "CloneSeq", 
  color_nodes_by = c("CloneSeq", "SampleID"), 
  color_scheme = c("plasma", "turbo"),
  size_nodes_by = 3, 
  output_dir = NULL, output_name = "Cluster 1")
```

# (Optional) K-means Clustering on TCR Sequences Encoded By Atchley Factors

**NOTE: This step applies to TCR CDR3 amino acid sequences only**

As an additional step in the analysis, we can take the TCR sequences from the full network in step 3 and use a deep learning algorithm with a trained encoder to encode each TCR sequence as a 30-dimensional numeric vector based on the Atchley-factor representations of its amino acids. This allows us to perform $K$-means clustering on the TCR sequences based on their numerically encoded values. 

We can then profile each sample based on how its TCR sequences are distributed among the $K$-means clusters: We compute the fraction of the sample's unique TCR sequences that belong to each cluster, yielding a $K$-dimensional vector representing the profile for the sample. We can then use heatmaps to compare these profiles across samples and to assess the correlation in profile between samples.

```{r, eval = FALSE}
atchley_results <- kmeansAtchley(
  data = all_clusters$node_data,
  amino_col = "CloneSeq", sample_col = "SampleID", group_col = "GroupID",
  k = 5, output_dir = file.path(data_dir, "assoc_clusters"), 
  return_output = TRUE)
```

The ultimate purpose of `kmeansAtchley()` is to produce the two heatmaps, which are generated using the `gplots::heatmap.2` function and saved to pdf. The numerically-encoded TCR values and the $K$-means sample profiles will be returned if `return_output = TRUE`.