---
title: "buildRepSeqNetwork"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{buildRepSeqNetwork}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center", 
  out.width = "100%",
  fig.width = 9, fig.height = 7
)
```


General network analysis on RepSeq data is performed using the `buildRepSeqNetwork()` function. This function does the following:

* Builds the network graph for the immune repertoire
* Computes desired network properties
* Prints a customized `ggraph` plot of the network graph
* Returns meta-data for the TCR/BCR (nodes) in the network, including biological as well as network properties
* Also returns the network `igraph` and adjacency matrix, as well as the `ggraph` plot object, for use in downstream analysis

In this vignette we demonstrate the function's basic usage.

### Load Data

For demonstration purposes, we simulate some toy data using built-in package functions.

```{r }
library(NAIR)
dir_out <- tempdir()
toy_data <- simulateToyData()
head(toy_data)
```


## Basic Usage

* The first argument specifies the data frame containing the rep-seq data, where each row corresponds to a single TCR/BCR clone (bulk data) or cell (single-cell data).
* The second argument specifies the column name or number of the data frame that contains the receptor sequences to be used as the basis of similarity between two cells or clones.

*For single-cell data, see [this vignette](dual_chain.html) on building cell-level networks based on joint similarity between alpha chain sequence and beta chain sequence.*

```{r}
output <- buildRepSeqNetwork(toy_data, seq_col = "CloneSeq", 
                             output_dir = dir_out)
```


The function returns a list containing the following items:
```{r}
names(output)
```

The item `node_data` is a data frame containing the same columns as the input data:

```{r}
names(output$node_data)
```

Only rows corresponding to nodes that remain in the network graph are included (those corresponding to the dropped isolated nodes have been removed):
```{r}
nrow(output$node_data)
```

Thus, this output data serves as biological meta-data for the nodes in the network graph, with each row corresponding to a node seen in the plot above.




## Network Properties


### Node-Level Network Properties

Use `node_stats = TRUE` to include node-level network properties. 

```{r, eval = FALSE}
# Node-level properties
output <- buildRepSeqNetwork(toy_data, "CloneSeq", node_stats = TRUE, 
                             output_dir = NULL)
```

The node data now contains node-level network properties in addition to the biological meta-data:
```{r, eval = FALSE}
names(output$node_data)
```


To choose which node-level network properties are computed, use the `node_stat_settings()` function, passing its output to the `stats_to_include` argument of `buildRepSeqNetwork()`...

```{r, eval = FALSE}
# example usage of node_stat_settings()
buildRepSeqNetwork(
  toy_data, "CloneSeq", node_stats = TRUE,
  stats_to_include = node_stat_settings(cluster_id = TRUE, closeness = FALSE))
```

...or include all network properties using either of the following methods.

```{r, eval = FALSE}
# the following two calls are equivalent:
buildRepSeqNetwork(
  toy_data, "CloneSeq", node_stats = TRUE, 
  stats_to_include = node_stat_settings(all_stats = TRUE))

buildRepSeqNetwork(
  toy_data, "CloneSeq", node_stats = TRUE, stats_to_include = "all")
```


### Cluster-Level Network Properties

```{r}
# Node-level and cluster-level properties
output <- buildRepSeqNetwork(toy_data, "CloneSeq", node_stats = TRUE, 
                             cluster_stats = TRUE, print_plots = FALSE,
                             output_dir = NULL)
```

The output list now contains an additional data frame for the cluster-level meta data:

```{r}
names(output)
head(output$cluster_data)
```

Each row of the cluster-level meta data corresponds to a single cluster in the network.

The `cluster_fun` argument controls the algorithm used to cluster the nodes; see `?addClusterMembership` for more details. The default value is `"cluster_fast_greedy"` (see `?igraph::cluster_fast_greedy`), but many other options exist.


## Network Settings

### Distance Function

By default, similarity between TCR/BCR sequences is based on the Hamming distance, i.e., the number of non-matching characters in two sequences of equal length. When sequence lengths do not match, we effectively extend the shorter sequence, with the additional terms treated as non-matching with those of the longer sequence.

Our package also supports the Levenshtein (edit) distance, which measures the minimum number of single-character edits (insertions, deletions and transformations) required to transform one sequence into the other. It can be used in place of the Hamming distance with the argument `dist_type = "levenshtein"`. Here we use the Levenshtein distance to build a network based off of similarity in the CDR-3 nucleotide sequences.

```{r, eval = FALSE}
# Network based on Levenshtein distance
buildRepSeqNetwork(toy_data, "cdr3_nt", dist_type = "levenshtein")
```

The Levenshtein distance innately applies to sequences of differing lengths and can correctly account for insertions and deletions, but is more computationally expensive, which could potentially pose challenges when working with very large data sets and using very long TCR/BCR sequences.


### Distance Cutoff

The distance function specified in the `dist_type` argument is used to model the similarity between TCR/BCR sequences. By default, two nodes in the network graph share an edge if their distance, as measured by this function, is at most 1. This cutoff value of 1 can be set to a different value, if desired, using the `dist_cutoff` argument.




## Other Arguments

### Input Filtering

* `min_seq_length` can be used to filter out TCR/BCR sequences by minimum length. Data rows with sequence lengths below this value will be dropped before computing the network graph. The default is 3.
* `drop_matches` can be used to filter out TCR/BCR sequences by content. It takes a character string or regular expression and checks each TCR/BCR sequence for a match; data rows with matches are dropped before computing the network graph. 

### Output 

* Output is saved to the current working directory by default; a directory can be specified using the `output_dir` argument. 
* The file type can be specified via the `output_type` argument: by default, each component of the returned list is saved to its own file, with the `output_name` argument used as a common file name prefix.
* For better compression, use a value of `"rds"` or `"rda"` for `output_type`, which will save the output list to a .rds or .rda file; the file will be named `output_name` with the appropriate file extension appended. 
* A pdf file of the graph plot(s) will be saved if `plots = TRUE`, regardless of the value of `output_type`. The dimensions (in inches) for the pdf can be adjusted using `plot_width` and `plot_height`, with the defaults being `12` and `10`.



### Visualization

Use  `print_plots = FALSE` to prevent plots from being printed to the R plotting window, and use `plots = FALSE` to prevent plots from being generated entirely. 

Numerous arguments exist that allow customization of the network visualization. See [this vignette](network_visualization.html) for further details.

