---
title: "Dual-Chain Network Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dual-Chain Network Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE, collapse = TRUE, comment = "#>", 
  fig.align = "center", fig.width = 10, fig.height = 10, out.width = "100%")
```

For single-cell data, cell-level network analysis can be performed based on joint similarity in alpha chain sequence and beta chain sequence.

We simulate some toy data to demonstrate the usage.

```{r}
library(NAIR)
dat <- simulateToyData(chains = 2)
head(dat)
```

The input data is assumed to have the following format:

* Each row corresponds to a unique cell
* The data contains separate columns for alpha chain sequence and beta chain sequence

Dual-chain network analysis can be performed using the `buildRepSeqNetwork` function (or `generateNetworkObjects` as a lower-level alternative) by supplying a length-2 vector to the `seq_col` argument:

* First entry should reference the column for alpha chain sequence
* Second entry should reference the column for beta chain sequence

```{r}
# Build network based on joint dual-chain similarity
network <- buildRepSeqNetwork(dat, seq_col = c("AlphaSeq", "BetaSeq"),
                              count_col = "UMIs",
                              node_stats = TRUE, stats_to_include = "all",
                              cluster_stats = TRUE, 
                              color_nodes_by = "SampleID",
                              size_nodes_by = "UMIs",
                              node_size_limits = c(0.5, 3),
                              print_plots = FALSE, output_dir = NULL)
```

We print the network graph plot with labels added for the largest two clusters:

```{r}
addClusterLabels(network$plots$SampleID, network, top_n_clusters = 2, size = 8)
```

The list returned by `buildRepSeqNetwork` contains the following items:

```{r}
names(network)
```

Notice that the list contains three adjacency matrices: `adjacency_matrix` corresponds to the network based on joint similarity in both chain sequences, while `adj_mat_a` corresponds to the network based only on similarity in the alpha-chain sequence (and similarly for `adj_mat_b`). 


The cluster-level data contains sequence-based cluster statistics for each of the alpha and beta chain sequences:

```{r}
head(network$cluster_data)
```


The remainder of the output and customization follows the [general case for the buildRepSeqNetwork function](buildRepSeqNetwork.html).