---
title: "Network Visualization"
output: 
  rmarkdown::html_vignette:
    number_sections: false
    tabset: true
vignette: >
  %\VignetteIndexEntry{Network Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center", 
  out.width = "100%",
  fig.width = 9, fig.height = 7
)
```

# Introduction

`buildRepSeqNetwork()` includes various arguments that facilitate customization of the network visualization. In this vignette, we provide an overview and demonstration of these arguments.


## Simulate Data for Demonstration

We simulate some toy data for demonstration.

We simulate data consisting of two samples with 100 observations each, for a total of 200 observations (rows).

```{r }
library(NAIR)
dir_out <- tempdir()
toy_data <- simulateToyData()
head(toy_data)
```

# Graph Elements {.tabset}

This section covers arguments that control aspects of the graph itself, such as the color and size of the nodes and the width of the edge connections.


## <a name="node_colors"></a> Node Colors {.tabset}

### Color Nodes According to Metadata

The nodes in the graph can be colored according to available metadata. This is accomplished using the `color_nodes_by` argument, which accepts a character string containing the name of a variable.

The variable can be any variable present in the input data, as well as any [node-level network properties](buildRepSeqNetwork.html#node_properties) that are computed for the network.

For example, when calling `buildRepSeqNetwork()`, we can color the nodes based on the `CloneCount` column of the input data by setting `color_nodes_by = "CloneCount"` as seen below.

```{r}
buildRepSeqNetwork(toy_data, 
                   seq_col = "CloneSeq", 
                   node_stats = TRUE,
                   color_nodes_by = "CloneCount", 
                   output_dir = NULL
)
```

If the `color_nodes_by` argument is left unspecified, `buildRepSeqNetwork()` will attempt to use an available variable to color the nodes. If a variable for abundance is provided using the `count_col` argument, this will be used. Otherwise the `cluster_id` variable will be used if it exists, followed by `degree`, then by other network properties. If no suitable options are available, the nodes will be left uncolored.

If the user does not wish the nodes to be colored dynamically, this can be specified by setting `color_nodes_by = NULL`. 


### Change Node Color Palette

A preset color palette can be used to color the nodes by providing the appropriate character string to the `color_scheme` argument. The argument accepts the following values:

* `"default"` for default `ggplot2` colors
* One of the following color scales from the `viridis` package. These scales are designed to maintain their perceptual uniformity (consistently across the scale, values close together appear similar, while values far apart appear distinct) when printed in grey scale or when viewed by individuals with common forms of color-blindness or color vision deficiency. More information, including images of the color scales, can be referenced in the "Introduction to the viridis color maps" vignette, linked on the [CRAN page for the viridis package](https://CRAN.R-project.org/package=viridis). For offline information, see the documentation for `ggraph::scale_color_viridis()`.
* `"magma"` (or `"A"`)
* `"inferno"` (or `"B"`)
* `"plasma"` (or `"C"`)
* `"viridis"` (or `"D"`)
* `"cividis"` (or `"E"`)
* `"rocket"` (or `"F"`)
* `"mako"` (or `"G"`)
* `"turbo"` (or `"H"`)
* Any of the above color scales with `"-1"` appended, which reverses the direction of the color scale (e.g., `color_scheme = "viridis-1"`)
* A discrete color palette from `grDevices::hcl.pals()` (these can only be used if the variable used to color the nodes is discrete)


Below we show an example of using the `plasma` color scheme from the `viridisLite` package, with the direction of the color scale reversed:

```{r}
# Using the "plasma" color scheme with reversed color gradient
buildRepSeqNetwork(toy_data, 
                   seq_col = "CloneSeq", 
                   node_stats = TRUE,
                   color_nodes_by = "transitivity", 
                   color_scheme = "plasma-1", 
                   output_dir = NULL
)
```

### Use Custom Color Palette

The color scales that are available through the `color_scheme` argument of `buildRepSeqNetwork()` are limited to a selection of preset color scales. If the user wishes to specify a custom color scheme for the nodes, this can be done by [modifying the plot after it is created](#modifying_plots). 


## <a name="node_size"></a> Node Size {.tabset}

### Uniform Node Size

By default, all nodes are drawn at a uniform size value of 0.5. This results in the nodes appearing very small when the plot is saved using the default pdf dimensions of 12 inches wide by 10 inches tall. This default behavior is intended to prevent nodes from overlapping and obscuring edges in larger networks. 

A different uniform node size can be specified by providing a numeric value to the `size_nodes_by` argument. Below, we set the node size to 1.5, which is three times as large as the default node size:

```{r}
buildRepSeqNetwork(toy_data, 
                   seq_col = "CloneSeq", 
                   node_stats = TRUE,
                   color_nodes_by = "transitivity", 
                   color_scheme = "plasma-1",
                   size_nodes_by = 1.5, 
                   output_dir = NULL
)
```



### Size Nodes According to Metadata

Rather than draw all nodes using a uniform node size, it is possible to for the nodes to be sized according to a variable within the node-level metadata. This is achieved by providing the column name of the variable to the `size_nodes_by` argument.

The minimum and maximum node sizes can be specified using the `node_size_limits` argument, which accepts a numeric vector of length two, where the first entry is the minimum node size and the second entry is the maximum node size.

```{r}
# Size nodes dynamically by network degree; use custom size range
buildRepSeqNetwork(toy_data, 
                   seq_col = "CloneSeq", 
                   node_stats = TRUE,
                   color_nodes_by = "transitivity", 
                   color_scheme = "plasma-1",
                   size_nodes_by = "degree", 
                   node_size_limits = c(0.5, 1.5), 
                   output_dir = NULL
)
```




## Edge width

By default, the edges in the plot are drawn with a width of 0.1.

A different width can be used for the edges by setting the `edge_width` argument to a different value.


# Text Elements {.tabset}

This section covers arguments that control text elements in the plot, such as titles, annotations and legends.



# <a name="labeling_clusters"></a> Labeling Clusters

If [cluster analysis is performed](buildRepSeqNetwork.html#cluster-analysis) when calling `buildRepSeqNetwork()`, the output will include data on the resulting clusters, with each cluster identified by a cluster ID.

In order to more easily reference these clusters within the visual plot of the network graph, it is possible to label the clusters in the plot with their cluster IDs. This must be done after calling `buildRepSeqNetwork()`, and is accomplished by using the `addClusterLabels()` to modify the plot contained in the output of `buildRepSeqNetwork()`. Note that `buildRepSeqNetwork()` returns a list, and one of the elements of this list is another list named `plots`, which contains all plots generated by `buildRepSeqNetwork()`.

`addClusterLabels()` has two primary arguments. The plot to be modified is provided to the `plot` argument. The entire output list returned by `buildRepSeqNetwork()` is provided to the `net` argument.

By default, only the 20 largest clusters by node count are labeled in order to preserve legibility. This number can be changed by providing a different value to the `top_n_clusters` argument. Instead of prioritizing the clusters to label based on their node count, a different variable within the cluster-level metadata can be used, so long as the variable is numeric. This is achieved by providing the column name of the variable to the `criterion` argument. The variable should be present in the `cluster_data` data frame contained in the output list of `buildRepSeqNetwork()`. Rather than prioritizing the clusters with the greatest values of this variable, those with the lowest values can instead be prioritized for labeling by setting `greatest_values = FALSE`.

The size of the cluster ID labels can be adjusted by providing a numeric value to the `size` argument (the default is 5), and their color can be adjusted by providing a valid character string to the `color` argument (the default is `"black"`).

`addClusterLabels()` assumes that the node-level metadata includes a variable named `cluster_id` that records the cluster membership of each node. This is the case when `buildRepSeqNetwork()` is called with cluster analysis enabled. If this variable has a name different from `cluster_id` (e.g., if columns are manually renamed), the correct variable name must be provided to the `cluster_id_col` argument.


```{r}
# Generate network, but don't print the graph plot yet
network <- 
  buildRepSeqNetwork(toy_data, 
                     seq_col = "CloneSeq", 
                     node_stats = TRUE, 
                     cluster_stats = TRUE,
                     color_nodes_by = "transitivity", 
                     color_scheme = "plasma-1",
                     size_nodes_by = "degree", 
                     node_size_limits = c(0.5, 1.5), 
                     print_plots = FALSE, 
                     output_dir = NULL
  )

# Add labels to the two largest clusters and print the plot
addClusterLabels(plot = network$plots$transitivity,
                 net = network,
                 top_n_clusters = 2,
                 criterion = "node_count" # (the default)
)
```



## Title and Subtitle

The `output_name` argument is used in the names of files saved by `buildRepSeqNetwork()`. By default, it is also used as the title for each plot generated by `buildRepSeqNetwork()`. The default value of `output_name` is `"MyRepSeqNetwork"`.

By default, the subtitle of each plot contains information about the settings used to construct the network, including the values of the `dist_type` and `dist_cutoff` arguments.

A custom plot title and subtitle can be specified using the `plot_title` and `plot_subtitle` arguments, respectively. Either element can be omitted from the plot by supplying a `NULL` value to the corresponding argument, as shown below.

```{r}
buildRepSeqNetwork(toy_data, 
                   seq_col = "CloneSeq", 
                   node_stats = TRUE,
                   color_nodes_by = "transitivity", 
                   color_scheme = "plasma-1",
                   size_nodes_by = "degree", 
                   node_size_limits = c(0.5, 2.5),
                   plot_title = NULL, 
                   plot_subtitle = NULL, 
                   output_dir = NULL
)
```



## Legends {.tabset}

### Excluding Size/Color Legend

By default, if the nodes are colored or sized dynamically according to a variable, a legend will be included in the plot showing the color scale and/or size scale. 

The color scale can be manually excluded from the legend by setting `color_legend = FALSE`. Similarly, setting `size_legend = FALSE` will exclude the size scale from the legend.

**Note:** An exception to the default behavior occurs when the variable used to color the nodes is discrete with more than 20 distinct values. In this case the color scale is automatically excluded from the legend in order to prevent it from taking up excessive space in the plot. The name of the variable used to color the nodes is then reported in the plot's subtitle for reference. If the user wishes to ignore this exception and force the color legend to be shown, they can do so by setting `color_legend = TRUE`. 


### Legend Titles

By default, the title shown above the color scale in the legend is the name of the variable used to color the nodes. A custom title can be specified by providing a character string to the `color_title` argument. Similarly, the default title for the size scale is the name of variable used to size the nodes, and a custom title can be specified using the `size_title` argument. Each of the `color_title` and `size_title` arguments can also be set to `NULL` in order to omit the respective title.

If a vector is provided to the `color_nodes_by` argument in order to [generate multiple plots](#multiple_plots), then the `color_title` will accept a character vector of matching length, where each entry is the title for the color legend in the corresponding plot.




# <a name="multiple_plots"></a> Generate Multiple Plots of a Network

The `buildRepSeqNetwork()` can generate multiple plots of the same network graph, with each plot coloring the nodes according to a different variable. This is accomplished by providing a vector of column names to the `color_nodes_by` argument instead of a single column name.

A different [color palette](#node_colors) can be used for each plot. This is achieved by providing a character vector to the `color_scheme` argument. This vector must have the same length as the vector provided to the `color_nodes_by` argument. The color palette specified by each entry will be applied to the corresponding plot. If, instead, a single value is provided to the `color_scheme` argument, the specified color palette will be applied to all of the plots.

```{r}
network <- 
  buildRepSeqNetwork(toy_data, 
                     seq_col = "CloneSeq", 
                     node_stats = TRUE, 
                     color_nodes_by = c("transitivity", "CloneCount"),
                     color_scheme = c("plasma-1", "default"),
                     size_nodes_by = "CloneCount", 
                     node_size_limits = c(0.1, 2.5),
                     output_dir = NULL
  )
```

Both plots are stored in the `plots` list, which is contained within the output list of `buildRepSeqNetwork()`. 

```{r}
names(network$plots)
```

Each plot is named according to the variable used to color the nodes.


# Downstream Tasks {.tabset}

This section covers functionality within the `NAIR` package that facilitates additional visual analysis of the results from `buildRepSeqNetwork()`.


## <a name="modifying_plots"></a> Modifying Plots After Creation {.tabset}


### Motivation

In some cases, the user may wish to make changes to a plot that was previously created by `buildRepSeqNetwork()`. 

For example, suppose we call `buildRepSeqNetwork()` in order to construct the network for our simulated data:

```{r}
output <- buildRepSeqNetwork(toy_data, "CloneSeq", 
                             node_stats = TRUE,
                             output_dir = NULL
)
```

Now suppose we wish to modify the plot seen above. We could call `buildRepSeqNetwork()` again, using the appropriate arguments to customize the plot as desired. However, this would involve reconstructing the network, which is unnecessary for our purposes and could take substantial time if the network is very large.

Another option is to directly modify the plot contained within the already-existing output of `buildRepSeqNetwork()`. 


### How to Modify Plots

The list that was returned by  `buildRepSeqNetwork()` contains a sublist named `plots`:

```{r}
names(output$plots)
```

Each plot created by `buildRepSeqNetwork()` is contained in this list, with each plot named according to the variable used to color the nodes. Here the list contains one plot named `degree`.

Each plot exists as an object of class `ggraph` from the `ggraph` package. The `ggraph` class inherits the structure of the `ggplot` class from the `ggplot2` package. Each plot can therefore be modified using `ggplot2` functions and syntax, for which [many excellent informational resources are available](https://ggplot2.tidyverse.org). 

The `ggraph` class extends the `ggplot` class through the addition of graph-specific geoms and layouts. The plots produced by `buildRepSeqNetwork()` use the geoms `geom_edge_link0()` for the edges and `geom_node_point()` for the nodes. 

As a demonstration below, we remove the plot title and subtitle, adjust the node size, color the nodes using the `CloneCount` variable and change the color scale. 

```{r}
library(ggraph)

# modify plot
modified_plot <- output$plots$degree + 
  # remove title and subtitle
  labs(title = NULL, subtitle = NULL) +  
  # adjust node size and node color variable
  geom_node_point(
    aes(color = output$node_data$CloneCount), size = 1.5) +
  # change legend title for node color
  guides(color = guide_legend(title = "Clone Count")) + 
  # change color scale
  scale_color_viridis(option = "plasma", direction = -1)

modified_plot
```

We can replace the original plot contained in the output of `buildRepSeqNetwork()` with the modified plot. If the variable used to color the nodes has been modified, we can also rename the plot accordingly.

```{r, eval = FALSE}
# replace original plot with modified plot
output$plots$degree <- modified_plot

# rename plot according to new node color variable
names(output$plots) <- "CloneCount"
```

Alternatively, we can keep the original plot and add the modified plot to the plot list contained in the existing output of `buildRepSeqNetwork()`.

```{r}
# add modified plot to list containing the original plot
output$plots$CloneCount <- modified_plot

names(output$plots)
```

The complete set of functions, geoms and aesthetics within the `ggraph` and `ggplot2` packages allow the user to customize plots beyond what is possible using only the arguments of `buildRepSeqNetwork()`.


### Custom Color Scales {.tabset}

The color scales that are available through the `color_scheme` argument of `buildRepSeqNetwork()` are limited to a selection of preset color scales. If the user wishes to manually customize the color scheme for the nodes, this can be done by modifying the plot after it is created.

#### For Continuous Variables

If the variable used to color the nodes is continuous, a custom color gradient can be specified using `scale_color_gradient()`, which allows the user to specify the color at each end of the gradient using the arguments `low` and `high`. Below, we use `buildRepSeqNetwork()` to generate a plot where the nodes are colored using the `CloneCount` variable. We then modify the plot to use a custom color gradient.

```{r}
# generate plot with nodes colored by clone count
output <- buildRepSeqNetwork(toy_data, "CloneSeq", 
                             color_nodes_by = "CloneCount", 
                             size_nodes_by = 1.5,
                             print_plots = FALSE, 
                             output_dir = NULL
)

# modify plot to use custom color gradient; replace original plot
output$plots$CloneCount <- output$plots$CloneCount +
  scale_color_gradient(low = "pink", high = "purple4")  

output$plots$CloneCount
```


#### For Discrete Variables

If the variable used to color the nodes is discrete, a custom color palette can be specified using the `scale_color_manual()` function, which allows the user to specify the color used for each value of the discrete variable. The argument `values` accepts a vector of colors whose length must match the number of unique values in the discrete variable. Below, we use `buildRepSeqNetwork()` to generate a plot where the nodes are colored using the `SampleID` variable. We then modify the plot to use a custom color palette.

```{r}
# generate plot with nodes colored by sample ID
output <- buildRepSeqNetwork(toy_data, "CloneSeq", 
                             color_nodes_by = "SampleID", 
                             size_nodes_by = 1.5,
                             print_plots = FALSE, 
                             output_dir = NULL
  )

# modify plot to use custom color palette
output$plots[[1]] + scale_color_manual(values = c("deepskyblue", "red2"))
```



## <a name="new_plots"></a> Generating New Plots

In some cases, rather than modifying an existing plot, the user may wish to generate an entirely new plot of the same network graph. For example, it may be the case that the user wants to keep their existing plots, but also wants additional plots with the nodes colored according to new variables not used in any of the current plots. 

While new plots can be generated by calling `buildRepSeqNetwork()` again, the user may wish to avoid reconstructing the same network, which is redundant and could be time-consuming for large networks.

Instead, new plots can be generated using `generateNetworkGraphPlots()` in combination with the existing output of `buildRepSeqNetwork()`. The function's first and second arguments accept the `igraph` and `node_data` objects, respectively, that are contained in the output list of `buildRepSeqNetwork()`.

The remaining arguments of `generateNetworkGraphPlots()` are used to customize the plot and behave identically to those of `buildRepSeqNetwork()` covered in this vignette.

`generateNetworkGraphPlots()` can be used to generate multiple plots in a similar fashion to `buildRepSeqNetwork()` by [supplying a vector of column references](#multiple_plots) to the `color_nodes_by` argument. 

The function returns a list containing one item for each plot. By default, the plots will also be printed to the plots pane in R. 

```{r}
# Generate new plots using previous output of buildRepSeqNetwork
new_plots <- 
  generateNetworkGraphPlots(igraph = output$igraph, 
                            data = output$node_data,
                            color_nodes_by = c("CloneCount", "CloneSeq"), 
                            color_scheme = c("default", "turbo"),
                            size_nodes_by = 1
  )
```

By default, plots generated using `generateNetworkGraphPlots()` do not have a title or subtitle. These can be set using the `plot_title` and `plot_subtitle` arguments, respectively.

The new plots can be added to the original list of plots by appending the list returned by `generateNetworkGraphPlots()`. 

```{r}
# add new plots to original list of plots
output$plots <- c(output$plots, new_plots)

names(output$plots)
```

<!-- The `plotNetworkGraph` function can be used to generate a single plot using the output of `buildRepSeqNetwork()`. The function's primary argument accepts the `igraph` object contained in the output list of `buildRepSeqNetwork()`.  -->

<!-- The remaining arguments of `plotNetworkGraph` are used to customize the plot and behave identically to those of `buildRepSeqNetwork()`, with one important exception that concerns the `color_nodes_by` and `size_nodes_by` arguments. In `buildRepSeqNetwork()`, each of these arguments accepts a column name or column index referencing the data column that contains the desired variable to be used. In `plotNetworkGraph`, however, each argument accepts the data column itself, rather than its name or index. -->

```{r eval=FALSE, include=FALSE}
plotNetworkGraph(igraph = output$igraph,
                 color_nodes_by = output$node_data$SampleID,
                 color_title = "Sample ID",
                 size_nodes_by = output$node_data$CloneCount,
                 size_title = "Clone Count",
                 node_size_limits = c(0.1, 2)
) 
```

<!-- By default, plots generated using `plotNetworkGraph` do not have a title or subtitle. These can be set using the `plot_title` and `plot_subtitle` arguments, respectively. -->


## Saving Plots

After making changes to the list of plots contained in the output of `buildRepSeqNetwork()`, such as replacing a plot with a modified version or adding new plots to the list, the user may wish to save the plots to a pdf file.

This can be conveniently accomplished by passing the list of plots to `saveNetworkPlots()` and specifying a file path for the pdf file using the `outfile` argument. All of the plots in the list will be saved to a single pdf.

```{r}
saveNetworkPlots(output$plots, 
                 outfile = file.path(dir_out, "plots.pdf")
)
```






```{r include=FALSE}
# clean up temp directory
file.remove(file.path(dir_out, "plots.pdf"))
```