---
title: "Network Visualization"
output: 
  rmarkdown::html_vignette:
    number_sections: false
    tabset: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center", 
  out.width = "100%",
  fig.width = 9, fig.height = 7
)
```

# Introduction

Customized visual plots of a network graph can be produced when calling `buildRepSeqNetwork()` and afterward using using `addPlots()`.


#### Simulate Data for Demonstration

```{r }
set.seed(42)
library(NAIR)
toy_data <- simulateToyData()
```

# Producing Plots

#### With `buildRepSeqNetwork()`/`buildNet()`

`buildRepSeqNetwork()` automatically generates a plot of the network graph by default:

```{r}
net <- buildNet(toy_data, "CloneSeq")
net$plots$uniform_color
```

#### <a name="generating-new-plots"></a><a name="new_plots"></a> With `addPlots()`

`addPlots()` can be used with the list returned by `buildRepSeqNetwork()` to generate additional plots of the network graph. 

```{r}
net <- addPlots(net, color_nodes_by = "SampleID")
net$plots$SampleID
```


### Reproducibility of Graph Layout

When creating the initial plot for a network, the coordinate layout of the nodes is generated pseudo-randomly. Other plots created in the same call to `buildRepSeqNetwork()` and subsequent calls to `addPlots()` will use the same graph layout as the initial plot (compare the previous two plots).

Different calls to `buildRepSeqNetwork()`, however, can produce plots of the same network with different layouts (compare the previous plot and next plot). For this reason, it is recommended to use `set.seed()` before calling `buildRepSeqNetwork()`. This allows the same graph layout to be reproduced across multiple executions of the same code in which the initial plots are generated.

```{r, eval = FALSE}
set.seed(42)
net <- buildNet(toy_data, "CloneSeq")
```



# <a name="node_colors"></a> Node Colors {.tabset}

#### Color Nodes According to Metadata

The nodes in the graph can be colored according to available metadata. The `color_nodes_by` parameter accepts a character string naming the variable to use. This can be any variable present in the [node metadata](https://mlizhangx.github.io/Network-Analysis-for-Repertoire-Sequencing-/articles/buildRepSeqNetwork.html#node-metadata).

```{r}
buildNet(toy_data, "CloneSeq", 
         color_nodes_by = "CloneCount",
         print_plots = TRUE
)
```

#### Node Color Scale

The `color_scheme` parameter accepts a character string naming a preset color scale to use when coloring the nodes. It accepts the following values:

* `"default"` for default `ggplot2` colors
* One of the following color scales from the [`viridis`](https://CRAN.R-project.org/package=viridis) package:
* `"magma"` (or `"A"`)
* `"inferno"` (or `"B"`)
* `"plasma"` (or `"C"`)
* `"viridis"` (or `"D"`)
* `"cividis"` (or `"E"`)
* `"rocket"` (or `"F"`)
* `"mako"` (or `"G"`)
* `"turbo"` (or `"H"`)
* Any of the above `viridis` color scales with `"-1"` appended (e.g., `"mako-1"`), which reverses the direction of the color scale
* If the variable is discrete, a color palette from `grDevices::hcl.pals()` (e.g., `"RdBu"`)

```{r}
net <- buildNet(toy_data, "CloneSeq", 
                node_stats = TRUE,
                color_nodes_by = "transitivity", 
                color_scheme = "plasma-1",
                print_plots = TRUE
)
```

#### <a name="modifying_plots"></a> Custom Color Scales {.tabset}

The color scales that are available through the `color_scheme` parameter are limited to a selection of preset color scales. A customized color scale can be applied to the plot after it is created.

If the variable used to color the nodes is continuous, a custom two-color gradient can be applied using `ggplot2::scale_color_gradient()`.

```{r}
net$plots$transitivity <- net$plots$transitivity +
  ggplot2::scale_color_gradient(low = "pink", high = "purple4")

net$plots$transitivity  
```

Similarly, `ggplot2::scale_color_gradient2()` can be used to apply a three-color diverging gradient.

If the variable used to color the nodes is discrete, a custom color palette can be applied using the `ggplot2::scale_color_manual()` function, which allows the user to specify the color used for each value. The parameter `values` accepts a vector of colors whose length matches the number of unique values.

```{r}
# Sample ID is a discrete variable
net <- addPlots(net, color_nodes_by = "SampleID")

net$plots$SampleID <- net$plots$SampleID + 
  ggplot2::scale_color_manual(values = c("deepskyblue", "red2"))

net$plots$SampleID
```

# <a name="node_size"></a> Node Size {.tabset}

By default, all nodes are drawn using a uniform size value of 0.5. This value is suitable for large networks with many nodes. 

The node size can be specified by providing a positive value to the `size_nodes_by` parameter. 

```{r}
net <- addPlots(net,
                color_nodes_by = "transitivity", 
                color_scheme = "plasma-1",
                size_nodes_by = 1.5,
                print_plots = TRUE
)
```

#### Size Nodes According to Metadata

The nodes can be sized dynamically according to a variable in the [node metadata](https://mlizhangx.github.io/Network-Analysis-for-Repertoire-Sequencing-/articles/buildRepSeqNetwork.html#node-metadata). The `size_nodes_by` parameter accepts a character string naming the variable.

The node size scale can be adjusted using the `node_size_limits` parameter, which accepts a numeric vector of length two specifying the minimum and maximum node size.

```{r}
net <- addPlots(net,
                color_nodes_by = "transitivity", 
                color_scheme = "plasma-1",
                size_nodes_by = "degree", 
                node_size_limits = c(0.1, 2.5),
                print_plots = TRUE
)
```



# Text Elements {.tabset}

This section covers parameters that control text elements in the plot, such as titles, annotations and legends.


### Title and Subtitle

The default title for plots created by `buildRepSeqNetwork()` is the argument value to the `output_name` parameter (`"MyRepSeqNetwork"` by default). Plots created by `addPlots()` have no default title.

For both `buildRepSeqNetwork()` and `addPlots()`, the default subtitle includes information about the network's construction, such as the values of `dist_type` and `dist_cutoff`.

A custom plot title and subtitle can be provided as character strings to the `plot_title` and `plot_subtitle` parameters. An argument value of `NULL` omits the corresponding element from the plot.

```{r, eval = FALSE}
net <- addPlots(
  net,
  plot_title = "Immune Repertoire Network by TCR Sequence Similarity",
  plot_subtitle = NULL
)
```



### Legends {.tabset}

#### Hiding Legends

By default, if the nodes are colored or sized dynamically according to a variable, a legend will be included in the plot showing the color scale and/or size scale. 

The color scale legend can be hidden using `color_legend = FALSE`. Similarly, `size_legend = FALSE` hides the size scale legend.

**Note:** When the variable used to color the nodes is discrete with more than 20 distinct values, the color scale is automatically excluded from the legend to prevent it from crowding the plot. The name of the node color variable is then appended to the plot's subtitle. To force the color legend to be shown, use `color_legend = TRUE`. 


#### Legend Titles

The default titles of the color and size scales are the names of the variables used to color and size the nodes. Custom titles can be provided as character strings to the `color_title` and `size_title` parameters. An argument value of `NULL` or `""` omits the title.

If a vector is provided to the `color_nodes_by` parameter in order to [generate multiple plots](#multiple_plots), then the `color_title` accepts a character vector of matching length, where each entry is the title for the color legend in the corresponding plot.



### <a name="labeling_clusters"></a> Labeling Clusters

After [performing cluster analysis](https://mlizhangx.github.io/Network-Analysis-for-Repertoire-Sequencing-/articles/cluster_analysis.html), it can be helpful to label clusters in the plot with their cluster IDs for reference, as shown [here](https://mlizhangx.github.io/Network-Analysis-for-Repertoire-Sequencing-/articles/cluster_analysis.html#labeling-clusters).


### Labeling Nodes

`addGraphLabels()` can be used to label individual nodes.



# <a name="multiple_plots"></a> Generating Multiple Plots

`buildRepSeqNetwork()` and `addPlots()` can generate multiple plots in a single call, with each plot coloring the nodes according to a different variable. The  `color_nodes_by` parameter accepts a character vector naming the variables to use. The `color_scheme` parameter will accept a character vector of matching length specifying the color scale for each plot, or a character string specifying a single color scale to use in all plots.

```{r}
net <- addPlots(net,
                color_nodes_by = c("transitivity", "CloneCount"),
                color_scheme = c("plasma-1", "default"),
                size_nodes_by = "CloneCount", 
                node_size_limits = c(0.1, 3),
                print_plots = TRUE
)
```




# Saving Plots

After adding or modifying plots in the list of network objects, the list of network objects can be saved using `saveNetwork()`, which also prints the plots to a PDF. `saveNetworkPlots()` saves the PDF only.

```{r, eval = FALSE}
saveNetwork(net, output_dir = dir_out)

saveNetworkPlots(net$plots, outfile = file.path(dir_out, "plots.pdf"))
```





