---
title: "Searching for Public TCR/BCR Clusters"
output: 
  rmarkdown::html_vignette:
    number_sections: false
    tabset: true
vignette: >
  %\VignetteIndexEntry{Searching for Public TCR/BCR Clusters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center", 
  out.width = "100%",
  fig.width = 9, fig.height = 7
)
```

# Introduction

The `NAIR` package includes a set of functions that facilitate searching for public TCR/BCR clusters across multiple samples of AIRR-seq data. 

In this context, a public cluster consists of similar TCR/BCR clones (e.g., those whose CDR3 amino acid sequences differ by at most one amino acid) that are shared across samples (e.g., across individuals or across time points for a single individual).

We first provide a brief conceptual overview, followed by a demonstration in which we explain the process in greater detail.


## Overview of Process

1. [**Filter the clusters within each sample's network**](#step1). For each sample, construct the repertoire network and use cluster analysis to partition the nodes into clusters.  Filter the data, keeping only those clusters with sufficient node count or clone count. 
2. [**Construct global network from the filtered data and perform clustering.**](#step2) Combine the filtered data from all samples into a single global network. Perform cluster analysis and assign membership to the global clusters. Assess sample representation within the clusters to identify public clusters.


## Simulate Data for Demonstration

We simulate some toy data for demonstration.

We simulate a total of 30 samples, each containing 30 observations. 

Some sequences are simulated with a tendency to appear in relatively few samples, while others are simulated with a tendency to appear in many samples. 

Each sample is saved in a separate file using the .rds file format. The files are named "`Sample1.rds`", "`Sample2.rds`", etc. The file path of their directory is saved to the R environment variable `dir_input_samples` for later reference.

```{r }
# Use temp dir
data_dir <- tempdir()

# Directory to store input files
dir_input_samples <- file.path(data_dir, "input_samples")
dir.create(dir_input_samples, showWarnings = FALSE)

samples <- 30
sample_size <- 30 # (seqs per sample)          

base_seqs <- c(
  "CASSIEGQLSTDTQYF", "CASSEEGQLSTDTQYF", "CASSSVETQYF",
  "CASSPEGQLSTDTQYF", "RASSLAGNTEAFF", "CASSHRGTDTQYF", "CASDAGVFQPQHF",
  "CASSLTSGYNEQFF", "CASSETGYNEQFF", "CASSLTGGNEQFF", "CASSYLTGYNEQFF",
  "CASSLTGNEQFF", "CASSLNGYNEQFF", "CASSFPWDGYGYTF", "CASTLARQGGELFF",
  "CASTLSRQGGELFF", "CSVELLPTGPLETSYNEQFF", "CSVELLPTGPSETSYNEQFF",
  "CVELLPTGPSETSYNEQFF", "CASLAGGRTQETQYF", "CASRLAGGRTQETQYF",
  "CASSLAGGRTETQYF", "CASSLAGGRTQETQYF", "CASSRLAGGRTQETQYF",
  "CASQYGGGNQPQHF", "CASSLGGGNQPQHF", "CASSNGGGNQPQHF", "CASSYGGGGNQPQHF",
  "CASSYGGGQPQHF", "CASSYKGGNQPQHF", "CASSYTGGGNQPQHF", 
  "CAWSSQETQYF", "CASSSPETQYF", "CASSGAYEQYF", "CSVDLGKGNNEQFF") 

# relative generation probabilities
pgen <- cbind(
  stats::toeplitz(0.6^(0:(sample_size - 1))),
  matrix(1, nrow = samples, ncol = length(base_seqs) - samples))

# Simulate the data
library(NAIR)
foo <- simulateToyData(    
  samples = samples, sample_size = sample_size,
  prefix_length = 1, prefix_chars = c("", ""),
  prefix_probs = cbind(rep(1, samples), rep(0, samples)),
  affixes = base_seqs, affix_probs = pgen, num_edits = 0,
  output_dir = dir_input_samples, no_return = FALSE)
```

The first few rows of the data for the first sample appear as follows:

```{r}
# View first few rows of data for sample 1
head(readRDS(file.path(dir_input_samples, "Sample1.rds")))
```


# <a name="step1"></a> Step 1: Filter Clusters Within Each Sample {.tabset}

First, we use the `findPublicClusters` function to perform network analysis on each sample individually and select clusters based on node count and clone count.

For each sample, the repertoire network is constructed and cluster analysis is used to partition the network into clusters. The clusters are then filtered according to node count and clone count based on user-specified criteria. The node-level and cluster-level metadata for the clusters that remain after filtering are saved as files to be used as inputs for [step 2](#step2).

Below, we explain the function's usage.


## Filter Settings {.tabset}

The `findPublicClusters` function has several parameters that control the criteria used to filter the nodes and clusters in each sample. These arguments are presented below.

### Top $n$ Clusters

Within each sample, the clusters are ranked by node count. The top $n$ highest-ranking clusters (those with the greatest node count) within each sample are automatically retained. The default value of $n$ is 20. A different value of $n$ can be specified using the `top_n_clusters` argument.

If more than one cluster is tied for $n$-th place in the ranking by node count, only one such cluster will be included in the top $n$ clusters. This ensures that no more than $n$ clusters are selected from each sample based on this criterion.

If fewer than $n$ clusters are present in the network for a sample, then all of the clusters will be retained.

### Minimum Node Count

In addition to retaining the top $n$ clusters from each sample, clusters that contain a sufficient number of nodes will also be retained. By default, any cluster containing at least ten nodes will be retained. This value can be adjusted using the `min_node_count` argument. For example, setting `min_clone_count = 30` will retain all clusters containing at least 30 nodes.


### Minimum Clone Count

In addition to the clusters retained based on node count, clusters with a sufficient aggregate clone count will also be retained. The aggregate clone count of a cluster is the sum of the clone counts across all nodes (clones) in the cluster. 

By default, any cluster with an aggregate clone count of at least 100 will be retained. This value can be adjusted using the `min_clone_count` argument. For example, setting `min_clone_count = 500` will retain all clusters with an aggregate clone count of at least 500.

Note that clusters are only selected according to this criterion if clone counts are provided. This is done using the `count_col` argument as explained in the **Specifying the Count Column** subsection.

### Sequence Length

When building the network for each sample, only clones whose receptor sequences are at least three characters in length will be included in the network. This minimum value for sequence length can be adjusted by setting the `min_seq_length` argument to a different value. Setting the value to `NULL` bypasses this check.

### Sequence Content

When building the network for each sample, all clones whose receptor sequences contain characters `*` or `_` will be omitted from the network. This can be changed using the `drop_matches` argument, which accommodates a character string or regular expression specifying the pattern of content to search for. The content of each clone's sequence is checked for a match to this pattern using the `grep` function from base R. If a match is found, the clone is omitted from the network. Setting the value to `NULL` bypasses this check. 

For details on how the pattern matching is performed, please refer to the base R documentation files for `regex` and `grep`.


## Input File Paths 

The main argument of the `findPublicClusters` function is the `file_list` argument, which accepts a vector containing file paths. Each path corresponds to a distinct AIRR-Seq data file representing an individual sample.

Below, we prepare the vector `input_files` to be provided to the `file_list` argument of `findPublicClusters`:

```{r}
# input files for step 1 (one per sample)
input_files <- file.path(dir_input_samples, paste0("Sample", 1:samples, ".rds"))
head(input_files)
```

## Input File Format 

The file format of the input files for `findPublicClusters` is specified using the `input_type` parameter. The supported formats include `"rds"`, `"rda"`, `"csv"`, as well as files that can be read using the `read.table` function, such as `"tsv"` and `"txt"`. 

For text formats such as `"csv"`, `"tsv"` and `"txt"`, users can specify the separation option by utilizing the `sep` argument. The default setting `sep = ""` accommodates all forms of white space, i.e., one or more spaces, tabs, newlines or carriage returns. In addition, it is important to note that the first line of the data is assumed to be the header by default. To disable this behavior and treat the first line as data, users must set the `header` parameter to `FALSE`.

For the `"rda"` (Rdata) format, the `data_symbols` argument is used to specify the name of each sample's AIRR-Seq data frame within the Rdata file (i.e., the name of the data frame as it will appear in the R environment when the Rdata file is loaded). The argument accepts a character string. This requires that each sample's data frame has the same name within its respective Rdata file. For example, if each sample's data frame is named `mydataframe` and saved using the command `save(mydataframe, file = myfilepath)`, where the file path `myfilepath` is different for each sample, then we would set `data_symbols = "mydataframe"`.

Our samples are stored in .rds files, so we use `input_type = "rds"`. 


## Specifying the Sequence Column 

Each clone's TCR/BCR sequence is used as the basis of similarity between clones in the network.

The `seq_col` argument is used to specify the column containing the clone sequences in the input data for each sample. The argument accepts either the column name or column index.

In our simulated data, the column containing the clone sequences is named `CloneSeq`.


## Specifying the Count Column

The clone count for each clone is used to filter the clusters within each sample.

The `count_col` argument is used to specify the column containing the clone counts in the input data for each sample. The argument accepts either the column name or column index.

In our simulated data, the column containing the clone sequences is named `CloneCount`.

## Assigning Custom Sample IDs

The sample ID for each clone will be included in the filtered data saved by `findPublicClusters`. This will allow us to distinguish the sample of origin for each clone after we combine the filtered data from all samples in [step 2](#step2).

By default, the samples are labeled numerically according to the order they appear in `file_list`. The `sample_ids` argument allows for custom sample IDs to be assigned, if desired. The argument accepts a vector of the same length as `file_list`. Each entry of `sample_ids` is assigned as a sample ID to the sample in the corresponding entry of `file_list`. 


## Network Settings Per Sample {.tabset}

The settings used to construct the network for each sample can be customized using the arguments below.


### Distance Function

The default method for measuring the similarity between TCR/BCR sequences is the Hamming distance. It calculates the number of differences between two sequences of the same length. If the sequences have different lengths, the shorter sequence is extended by adding non-matching characters to make it the same length as the longer sequence.

The Levenshtein distance can be used as an alternative measurement to determine the similarity between sequences. It calculates the minimum number of single-character edits (insertions, deletions and transformations) needed to transform one sequence into the other. This method is particularly useful for comparing sequences of different lengths and can account for insertions and deletions. When constructing a network based on the similarity of CDR-3 nucleotide sequences, it is preferable to use the Levenshtein distance instead of the default Hamming distance by setting the argument `dist_type = "levenshtein"`. However, the Levenshtein distance requires significantly more computational time than the Hamming distance, which may be challenging when working with large data sets having long TCR/BCR sequences.


### Distance Cutoff

The distance function specified in the `dist_type` argument (Hamming distance by default) is used to quantify the similarity between TCR/BCR sequences. The chosen distance measurement determines the distance between two nodes in the network graph. 

By default, two nodes in the  graph are connected by an edge if their distance is at most 1. However, if users want to adjust this cutoff, the `dist_cutoff` argument can be set to a different value. 

For example, if `dist_cutoff = 2`, then two nodes will be connected by an edge if their distance is at most 2. The cutoff value controls the stringency of the network construction and affects the number and density of edges in the network.


### Keep/Remove Isolated Nodes 

By default, if a given node has no edges connecting it to any other nodes in the network, it will be removed from the network graph and will not be included in the output. 

If desired, all nodes can be kept in the network, including those that do not have any edge connections to other nodes. This is accomplished by setting the `drop_isolated_nodes` argument to `FALSE`.


## Clustering Algorithm

By default, clustering is performed using the `cluster_fast_greedy` algorithm from the `igraph` package. Other clustering algorithms from the `igraph` package can be used instead of the default algorithm. The algorithm is specified using the `cluster_fun` argument, which accepts one of the following functions:

* `cluster_fast_greedy` 
* `cluster_edge_betweenness` 
* `cluster_fluid_communities` 
* `cluster_infomap` 
* `cluster_label_prop` 
* `cluster_leading_eigen` 
* `cluster_leiden` 
* `cluster_louvain` 
* `cluster_optimal` 
* `cluster_spinglass` 
* `cluster_walktrap`

For example, setting `cluster_fun = cluster_leiden` performs clustering using the `cluster_leiden` algorithm.

For more information about a particular algorithm, users can refer to its help documentation file. For example, the command `?igraph::cluster_fast_greedy` loads the documentation file for the `cluster_fast_greedy` algorithm.


## Output Settings {.tabset}

The `findPublicClusters` function does not return any direct output. Instead, it saves the network metadata for the selected clusters to files that will be used as inputs in step 2. 

For each sample, `findPublicClusters` saves two files, one containing the node-specific network metadata and the other containing the cluster-specific network metadata. The data for each sample is saved after filtering the clusters, and thus includes only the clusters selected based on the specified filter settings.

### Subsetting Columns From the Input Data

The node-specific network metadata for each sample includes all of the data columns that were present in the sample's original input data frame. 

In some cases, the user may not wish for all columns to be included. In this case, the user can specify the columns to be included using the `subset_cols` argument, which accepts a vector containing either the column names or the column indices of the columns to be kept. 

The column specified by the `seq_col` argument will automatically be included regardless of the value of `subset_cols`.


### Output Directory

The file path for the output directory is specified using the `output_dir` argument. The output directory specified by the `output_dir` argument will be created if it does not already exist. 

Within the output directory, two subdirectories are created. One subdirectory, named `node_meta_data`, contains the node-level data files for each sample, and the other, named `cluster_meta_data`, contains the cluster-level data files. Within each of these two subdirectories, the file for each sample is named according to its sample ID as specified by the `sample_ids` argument (see the **Assigning Custom Sample IDs** subsection above). 

### Output File Format

By default, each file is saved as a RDS file. This can be changed using the `output_type` argument. Other valid options include `"rda"` and `"csv"`. For the `"rda"` (Rdata) format, each data frame containing node-level data will be named `ndat` in the R environment when its Rdata file is loaded, while each data frame containing cluster-level data will be named `cdat`.


### Saving Unfiltered Network Data

By default, the `findPublicClusters` function saves the network metadata only for the clusters selected based on the specified filter settings.

If desired, the network metadata for each sample's full network can also be saved prior to filtering the clusters. This is done by by providing a file path to the `output_dir_unfiltered` argument, which specifies a separate output directory for the full network data. Each sample's full network data prior to filtering the clusters will then be saved to the directory specified by the `output_dir_unfiltered` argument. This data is saved separately from, and in addition to, the default data that is saved after filtering the clusters. 

Note that the sequence-based filter settings specified by `min_seq_length` and `drop_chars` still apply to each sample's full network, since the network is only constructed after applying these filters. The full, pre-filtered network refers to the network that contains all of the sample's clusters, i.e., before the public clusters are identified and other clusters are removed.

The full network data for each sample includes node-level metadata, cluster-level metadata, the `igraph` edge list, the network adjacency matrix, as well as any plots generated (see next subsection). By default, the R objects for these files are saved into a single RDS file whose file name (excluding the .rds file extension) is the sample ID (as specified by the `sample_ids` argument). The RDA file format can be used instead of RDS by setting `output_type_unfiltered = "rda"`. If desired, each R object can be saved to a separate file by setting `output_type_unfiltered = "individual"`. This saves each object according to the [default file format used by `buildRepSeqNetwork`](buildRepSeqNetwork.html#output_settings). When saving objects individually, the sample ID is used as the common file name prefix for the files from each sample.


## (Optional) Network Visualization Per Sample

By default, `findPublicClusters` does not produce visual plots when constructing the network for each sample. Instead, visualization occurs after combining the data from all samples into a single network in [step 2](#step2).

If desired, visual plots of the full network for each sample (prior to filtering the clusters) can be produced by setting `plots = TRUE`. The plots can be printed in R by additionally setting `print_plots = TRUE`. Furthermore, if the user has specified to save the full network data for each sample (prior to filtering the clusters) by providing a file path to the `output_type_unfiltered` argument, these plots will be saved along with the rest of the full network data for each sample.

Note that the plots are not saved along with the default data that is saved after filtering the clusters. Therefore, if no file path is provided to the `output_type_unfiltered` argument, the plots will not be saved at all. If, in addition, the argument `print_plots` is set to `FALSE` (the default), setting `plots = TRUE` will have no effect.

If the user sets `plots` to `TRUE`, then by default, the plot for each sample will color the nodes according to their cluster membership. If desired, a different variable can be used to color the nodes. This is done by specifying the column for the variable to the `color_nodes_by` argument, which accepts a column name or column index. 

[This vignette](network_visualization.html) details additional arguments that can be used to further customize the visual plot for each sample.


## Example Function Call

We execute the `findPublicClusters` function using the input we prepared earlier for the `file_list` argument.

The file path of our output directory is saved to the R variable named `dir_filtered_samples` for later reference.

```{r}
# 1. Filter Clusters Within Each Sample
dir_filtered_samples <- file.path(data_dir, "filtered_samples")
findPublicClusters(
  file_list = input_files, input_type = "rds",
  sample_ids = paste0("Sample", 1:samples),
  seq_col = "CloneSeq", count_col = "CloneCount",
  min_seq_length = NULL, drop_matches = NULL,
  top_n_clusters = 3, min_node_count = 5, min_clone_count = 15000,
  output_dir = dir_filtered_samples)
```




# <a name="step2"></a> Step 2: Global Network of Public Clusters {.tabset}

Next, we use the `buildPublicClusterNetwork` function to combine the filtered data from all samples into a single global network and perform clustering analysis.  

## Input File Paths

The files created by `findPublicClusters` in the previous step contain the filtered data for each sample. As detailed earlier, these files are located in two separate subdirectories, one containing the files for the node-level metadata and the other containing the files for the cluster-level metadata. 

For this step, we require only the files containing the node-level metadata. These files are provided to `buildPublicClusterNetwork` by supplying a character vector of file paths to the `file_list` argument. We create this vector below.


```{r}
# Node-level metadata for each sample's filtered clusters
dir_filtered_samples_node <- file.path(dir_filtered_samples, "node_meta_data")
files_filtered_samples_node <- list.files(dir_filtered_samples_node, 
                                          full.names = TRUE)
head(files_filtered_samples_node)
```



## Input File Format 

By default, `findPublicClusters` saves its output files using the RDS file format. By default,  `buildPublicClusterNetwork` assumes that the input files are in the RDS file format. 

If the files created by `findPublicClusters` were saved using a different file format, the `input_type` argument can be used to specify the format used. This argument has the same behavior as seen in `findPublicClusters`.

For the `"rda"` (Rdata) file format, the `data_symbols` argument is used to specify the name of each sample's node-level data frame within its respective Rdata file. If `findPublicClusters` was called with `output_type = "rda"`, then the node-level data frame for each sample will be named `ndat`. Accordingly, the default value of `data_symbols` is `ndat`, so the user should not need to manually specify the value of `data_symbols` unless the names within the files are changed.

## Specifying the Sequence and Count Columns

The `seq_col` and `count_col` arguments are used to specify the columns containing the receptor sequences and clone counts, respectively, in the same manner seen earlier for `findPublicClusters`.

The clone counts are used to compute some of the cluster-level properties, which will be omitted if `count_col` is left unspecified.


## Network Settings

The settings used to construct the network can be customized [using the same arguments from `findPublicClusters`](buildRepSeqNetwork.html#network_settings), shown below with their default values:

* `dist_type = "hamming"` 
* `dist_cutoff = 1`
* `drop_isolated_nodes = FALSE`

Refer to [this vignette](buildRepSeqNetwork.html#network_settings) for details on their meaning and behavior.


## Clustering Algorithm 

After constructing the global network, `buildPublicClusterNetwork` performs cluster analysis on the network nodes, partitioning the global network graph into densely-connected subgraphs. These global clusters can contain nodes from different samples.

By default, clustering is performed using the `cluster_fast_greedy` algorithm from the `igraph` package. Other clustering algorithms from the `igraph` package can be used instead of the default algorithm. The algorithm is specified using the `cluster_fun` argument, which accepts one of the following functions:

* `cluster_fast_greedy` 
* `cluster_edge_betweenness` 
* `cluster_fluid_communities` 
* `cluster_infomap` 
* `cluster_label_prop` 
* `cluster_leading_eigen` 
* `cluster_leiden` 
* `cluster_louvain` 
* `cluster_optimal` 
* `cluster_spinglass` 
* `cluster_walktrap`

For example, setting `cluster_fun = cluster_leiden` performs clustering using the `cluster_leiden` algorithm.

For more information about a particular algorithm, users can refer to its help documentation file. For example, the command `?igraph::cluster_fast_greedy` loads the documentation file for the `cluster_fast_greedy` algorithm.


## Customization of Visual Plot

By default, the network graph plot produced by `buildPublicClusterNetwork` colors the nodes according to sample ID. This can assist the user in identifying the public clusters. If desired, a different variable can be used to color the nodes. This is done by specifying the column for the variable to the `color_nodes_by` argument, which accepts a column name or column index. 

[This vignette](network_visualization.html) details additional arguments that can be used to further customize the visual plot.


## Output Settings {.tabset}

The output returned by `buildPublicClusterNetwork` follows the [same format as the output of `buildRepSeqNetwork`](buildRepSeqNetwork.html#output_settings). The function returns a list containing the node-level and cluster-level metadata for the global network, as well as any plots generated, in addition to the network adjacency matrix and the `igraph` network edge list.

### Output Directory

By default, a subdirectory named `public_clusters` is created within the current working directory, and the contents of the list returned by `buildPublicClusterNetwork` are saved to this subdirectory.

The output can be saved to a different directory by providing a file path to the `output_dir` argument. 

The user can also specify `output_dir = NULL` in order to prevent the output from being saved.

### Output File Format

By default, each element of the list returned by `buildPublicClusterNetwork` is saved as an individual file. The file formats are the same default file formats used by [`buildRepSeqNetwork`](buildRepSeqNetwork.html#output_settings). In particular, the node-level and cluster-level metadata are saved as csv files. 

Alternatively, the user can save the entire output list to a single compressed rds or rda file by setting `output_type = "rds"` or `output_type = "rda"`, respectively.

### PDF of Plots

* Regardless of the value of `output_type`, a separate pdf file containing the graph plot(s) is created in `output_dir`. The dimensions (in inches) for the pdf can be adjusted using the `plot_width` and `plot_height` arguments, with the defaults being `12` and `10`, respectively. The pdf file is not created if no plots are generated (i.e., if the argument `plots` is set to `FALSE`).
* **Note:** the `ggraph` object for each plot is only saved if the user sets `output_type = "rds"` or `output_type = "rda"`. Using one of these settings is recommended used if the user wishes to modify any plots in the future. Note, however, that plots can always be re-generated from the node-level meta data using the `generateGraphPlot` function as described [here](network_visualization.html#new_plots).


### Output File Name(s)

* By default, the name of each saved file starts with `PublicClusterNetwork`. This prefix can be changed by providing a character string to the `output_name` argument.
* If `output_type` is set to `"rds"` or `"rda"`, then the name of the file will be the value of the `output_name` argument followed by the appropriate file extension (either `.rds` or `.rda`).
* The name of the pdf file for the graph plot will be the value of the `output_name` argument followed by the `.pdf` file extension.



## Example Call and Output {.tabset}

We execute the `buildPublicClusterNetwork` function using the input we prepared earlier for the `file_list` argument:

```{r}
dir_out <- file.path(data_dir, "public_clusters")

# Collect clones from all public clusters and perform network analysis
public_clusters <- buildPublicClusterNetwork(
  file_list = files_filtered_samples_node,
  seq_col = "CloneSeq", count_col = "CloneCount",
  size_nodes_by = 1,
  output_dir = dir_out)
```

The default files saved by `buildPublicClusterNetwork` have the following names:

```{r}
# Files saved by buildPublicClusterNetwork
list.files(dir_out)
```


The elements of the output list returned by `buildPublicClusterNetwork` have the following names:

```{r}
# output returned by buildPublicClusterNetwork
names(public_clusters)
```

Below, we describe each element of the above list.

### `node_data`

The `node_data` data frame contained in the output list contains one row for each node in the network, with the following variables:

```{r}
# variables in the node-level metadata
names(public_clusters$node_data)
```

Notice that by default, all variables that were present in each sample's original data, such as `"CloneFrequency"` and `"CloneCount"`, are automatically carried over into this data.

Many node-level network properties are present in the data. Some pertain to the network for the individual sample from which each node originated, i.e., the networks constructed in [step 1](#step1). This includes all variables that begin with `"SampleLevel"`, such as `SampleLevelNetworkDegree`, as well as the variable `ClusterIDInSample`.

The variables that begin with `"Public"`, such as `PublicNetworkDegree`, pertain to the global network. In particular, `ClusterIDPublic` indicates the ID of the global cluster to which each node belongs.


### `cluster_data`

The `cluster_data` data frame contained in the output list contains one row for each cluster in the network, with the following variables representing the cluster-level network properties:

```{r}
# variables in the node-level metadata
names(public_clusters$cluster_data)
```


### `plots`

The item `plots` is a list containing each of the plot(s) created by `buildPublicClusterNetwork`. Each plot is an object of class `ggraph` (which inherits from the `ggplot` class), and is named according to the variable used to color the nodes.

```{r}
names(public_clusters$plots)
```
```{r}
class(public_clusters$plots$SampleID)
```

The plots can be modified after their creation as described in [this vignette](network_visualization.html#modifying_plots). 

New plots can be generated from the output as described [here](network_visualization.html#new_plots).


### `igraph`

The item `igraph` is an object of class `igraph`, which contains the list of nodes and edges that comprise the network graph. 

It can be used to [generate additional plots of the network graph](network_visualization.html#new_plots).


### `adjacency_matrix`

The item `adjacency_matrix` is an object of class `dgCMatrix`, which is a sparse matrix format from the `Matrix` package. 

```{r}
class(public_clusters$adjacency_matrix)
```

The network adjacency matrix encodes the graph (i.e., the set of nodes and edges) as a matrix. It is symmetric, with each row (equivalently, each column) corresponding to a distinct network node. The network nodes follow the same order in the rows as they do in the columns. Each column name is the receptor sequence corresponding to the node for that column. Each row name is the row index of the original input data that corresponds to the node for that row.

```{r}
# First 5 rows and columns of the network adjacency matrix
head(public_clusters$adjacency_matrix, c(5, 5))
```

Each matrix entry has a value of 0 or 1, with 1 indicating an edge connecting the node for that row to the node for that column, and 0 indicating no edge connection. Although the diagonal entries are 1, their values are not meaningful, since an edge is not drawn from a given node to itself.

Note that the rows (nodes) in the adjacency matrix correspond to the rows (nodes) in the `node_data` data frame. Both have the same number of rows and the same row names.


## Labeling the Global Clusters

In the plot generated by `buildPublicClusterNetwork`, there appear to be six clusters with more than three samples represented. 

In order to reference these clusters within the data, we can [label the six largest clusters in the plot with their cluster IDs](network_visualization.html#labeling_clusters) using the `addClusterLabels` function. Note that within the node-level metadata, the global cluster ID is stored in the variable `ClusterIDPublic`, so we must provide this column name to the `cluster_id_col` argument.

```{r}
# Modify plot to add labels to the clusters
public_clusters$plots[[1]] <- 
  addClusterLabels(
    plot = public_clusters$plots[[1]],
    net = public_clusters,
    top_n_clusters = 6,
    cluster_id_col = "ClusterIDPublic",
    size = 7
  )

# View modified plot
public_clusters$plots[[1]]
```


## Focusing on Individual Clusters of Interest

If we wish to focus on a particular cluster of interest within the global network, we can build a network exclusively using the clones from that cluster. This is accomplished using the `buildRepSeqNetwork` function, where we subset our data according to the value of the `ClusterIDPublic` column, which contains the global cluster IDs.

Below, we focus on the first cluster, which in this case is also the largest cluster by node count. In the plot, we color each node according to the receptor sequence of its corresponding clone. 

```{r}
# focus on cluster 1
buildRepSeqNetwork(
  data = 
    public_clusters$node_data[public_clusters$node_data$ClusterIDPublic == 1, ],
  seq_col = "CloneSeq", 
  color_nodes_by = "CloneSeq", color_scheme = "plasma", 
  size_nodes_by = 3, 
  output_dir = NULL, output_name = "Cluster 1")
```

Below, we do the same for the sixth largest cluster:

```{r}
# focus on cluster 6
buildRepSeqNetwork(
  data = 
    public_clusters$node_data[public_clusters$node_data$ClusterIDPublic == 6, ],
  seq_col = "CloneSeq", 
  color_nodes_by = "CloneSeq", color_scheme = "plasma", 
  size_nodes_by = 3, 
  output_dir = NULL, output_name = "Cluster 6")
```

